Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-11-09 paper/main.tex
try to uniformize notations

A clean theoretical framework for compositional bounds sometimes
even allow to get better asymptotic bounds.  During our experiments
we found an example of this situation in some code from the cBench
benchmark suite.
%
\begin{lstlisting}
for (;;) {
    do { l++; tick(1); } while (l < h && nondet());
    do { h--; tick(1); } while (h > l && nondet());
    if (h <= l) break;
    tick(1); /* swap two elements */ }
\end{lstlisting}
%
The above snippet is the inner loop of a Quicksort implementation, more
precisely, it is the partitioning part of the algorithm.  This partition
loop has linear complexity, and feeding into our analysis gives the
correct and tight worst-case bound $2+3|[l,h]|$.  However, all available
tools for C will derive a quadratic bound instead.  With the amortized
analysis approach in mind, it is clear that the three loops get their
potential from the same interval $[l,h]$.  But on the other hand, current
techniques will try to find one linear ranking-function for each loop
and combine them multiplicatively or additively, resulting in a loose
quadratic bound.
%
\jan{Did you test the last example with other tools? What are the
  bounds they get? State the bounds here.}

@@

\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant } \{ \code{na} = \#_1(\code{a})} \}$
while (k > 0) {
  x=0;
  while (x < N && a[x] == 1 _&& na > 0_) {
    a[x]=0; _na--;_
    tick(1);
    x++; }
  if (x < N) {
    a[x]=1; _na++;_
    tick(1); }
  k--;
}
\end{lstlisting}
\end{minipage}
\caption{Example of assisted bound derivation using logical state.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.  The bound derived automatically is
  $2|[0,\code{k}]| + |[0,\code{na}]|$.
  }
\label{fig:xmplincaux}
\end{figure}

\begin{figure}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant }
  \{ \code{lg} > \log_2(\code h-\code l) \} }$
bsearch(x,l,h_,lg_) {
  if (h-l > 1 _&& lg > 0_) {
    m = l + (h-l)/2;
    _lg--;_ if (a[m]>x) h=m; else l=m;
    tick($M_\code{bsearch}$);
    l = bsearch(x,l,h_,lg_);
    tick($-M_\code{bsearch}$);
  } else return l;
}
\end{lstlisting}
\caption{Example of assisted bound derivation using logical state.
  We use the tick metric to model the stack consumption of the program.
  Before a function call to \code{bsearch}, a tick statement pays for
  the space used by stack frame of the function.  The space is recoverd
  after the call.  We write $\log_2(x)$ for the integer part of
  logarithm of $x$ in base 2.
  The bound derived automatically is $|[0,\code{lg}]|$.
  }
\label{fig:xmplbsaux}
\end{figure}

@@

This idea is illustrated by the program on \pref{fig:xmplincaux}.
Parts of the code underlined are annotations that were added
to the original program text.  The top-level loop increments
a binary counter \code{k} times.  A naive analysis of
the algorithm yields the quadratic bound $\code{k}\cdot\code{N}$.
However, the algorithm is in fact linear and its cost is bounded
by $2\code{k} + \#_1(\code{a})$ where $\#_1(\code{a})$ denotes
the number of one entries in the array \code{a}.  Since this
number depends on the heap contents, no tool available
for C is able to derive the linear bound.  However, it can be inferred
by our automated tool if a logical variable \code{na} is introduced.
This logical variable is a reification, in the program, of the
number $\#_1(\code{a})$.  For example, on line 5 of the example
we are setting \code{a[x]} to 0 and because of the condition we
know that this array entry was 1.  To reflect this change on
$\#_1(\code{a})$, the logical variable \code{na} is decremented.
Similarly, on line 9, an array entry which was 0 becomes 1, so
\code{na} is incremented.  To complete the step 2 of the systematic
procedure described above, we must show that the extra test
\code{na > 0} at line 4 does not influence the loop behavior.  It
is done by proving inductively that $\code{na} = \#_1(\code{a})$
and remarking that since $\code{a[x]} = 1$ is true, we must have
$\#_1(\code{a}) > 0$, thus \code{na > 0} in the test. %  So the
% test is in fact semantically equivalent to \code{true}.

Another simple example is given on \pref{fig:xmplbsaux} where
a logarithmic bound on the stack consumption of a binary search
program is proved using logical variable annotations.  Once again,
annotations are underlined in the program text.  In this example,
to ease the proof of equivalence between the annotated program
and the original one, we use the inequality $\code{lg} >
\log_2(\code h-\code l)$ as invariant.  This trick allows a simpler
proof because, when working with integer arithmetic, it is not always
the case that $\log_2(x-x/2) = \log_2(x)-1$ (take 3).

>>>>

A clean theoretical framework for compositional bounds sometimes
even allow to get better asymptotic bounds.  During our experiments
we found an example of this situation in some code from the cBench
benchmark suite.
%
\begin{lstlisting}
for (;;) {
    do { l++; tick(1); } while (l < h && nondet());
    do { h--; tick(1); } while (h > l && nondet());
    if (h <= l) break;
    tick(1); /* swap two elements */ }
\end{lstlisting}
%
The above snippet is the inner loop of a Quicksort implementation, more
precisely, it is the partitioning part of the algorithm.  This partition
loop has linear complexity, and feeding into our analysis gives the
correct and tight worst-case bound $2+3|[l,h]|$.  With the amortized
analysis approach in mind, it is clear that the three loops get their
potential from the same interval $[l,h]$.  However, KoAT fails to
prove the termination of the loop and LOOPUS derives the quadratic bound
$(h-l-1)^2$.  Following the classic technique, these tools try to find one
linear ranking-function for each loop and combine them multiplicatively
or additively, resulting in a loose quadratic bound or no bound at all.

@@

\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
while (k > 0) {
  x=0;
  while (x < N && a[x] == 1 _&& na > 0_) {
    a[x]=0; _na--;_
    tick(1);
    x++; }
  if (x < N) {
    a[x]=1; _na++;_
    tick(1); }
  k--;
}
\end{lstlisting}
\end{minipage}
\caption{Example of assisted bound derivation using logical state.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.  The bound derived automatically is
  $2|[0,k]| + |[0,na]|$.
  }
\label{fig:xmplincaux}
\end{figure}

\begin{figure}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant }
  \{ lg > \log_2(h - l) \} }$
bsearch(x,l,h_,lg_) {
  if (h-l > 1 _&& lg > 0_) {
    m = l + (h-l)/2;
    _lg--;_ if (a[m]>x) h=m; else l=m;
    tick($M_\code{bsearch}$);
    l = bsearch(x,l,h_,lg_);
    tick($-M_\code{bsearch}$);
  } else return l;
}
\end{lstlisting}
\caption{Example of assisted bound derivation using logical state.
  We use the tick metric to model the stack consumption of the program.
  Before a function call to \code{bsearch}, a tick statement pays for
  the space used by stack frame of the function.  The space is recoverd
  after the call.  We write $\log_2(x)$ for the integer part of
  logarithm of $x$ in base 2.
  The bound derived automatically is $|[0,lg]|$.
  }
\label{fig:xmplbsaux}
\end{figure}

@@

This idea is illustrated by the program on \pref{fig:xmplincaux}.
Parts of the code underlined are annotations that were added
to the original program text.  The top-level loop increments
a binary counter $k$ times.  A naive analysis of
the algorithm yields the quadratic bound $k \cdot N$.
However, the algorithm is in fact linear and its cost is bounded
by $2k + \#_1(a)$ where $\#_1(a)$ denotes
the number of one entries in the array $a$.  Since this
number depends on the heap contents, no tool available
for C is able to derive the linear bound.  However, it can be inferred
by our automated tool if a logical variable $na$ is introduced.
This logical variable is a reification, in the program, of the
number $\#_1(a)$.  For example, on line 5 of the example
we are setting \lstinline{a[x]} to 0 and because of the condition we
know that this array entry was 1.  To reflect this change on
$\#_1(a)$, the logical variable $na$ is decremented.
Similarly, on line 9, an array entry which was 0 becomes 1, so
$na$ is incremented.  To complete the step 2 of the systematic
procedure described above, we must show that the extra test
\lstinline{na > 0} at line 4 does not influence the loop behavior.  It
is done by proving inductively that $na = \#_1(a)$
and remarking that since \lstinline{a[x] == 1} is true, we must have
$\#_1(a) > 0$, thus \lstinline{na > 0} is true in the test. %  So the
% test is in fact semantically equivalent to \lstinline{true}.

Another simple example is given on \pref{fig:xmplbsaux} where
a logarithmic bound on the stack consumption of a binary search
program is proved using logical variable annotations.  Once again,
annotations are underlined in the program text.  In this example,
to ease the proof of equivalence between the annotated program
and the original one, we use the inequality $lg >
\log_2(h-l)$ as invariant.  This trick allows a simpler
proof because, when working with integer arithmetic, it is not always
the case that $\log_2(x-x/2) = \log_2(x)-1$ (take 3).
